---
title: "longitudinal_switching_analysis"
output: html_document
date: '2024-08-27'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(data.table)
library(ggplot2)
library(dplyr)
library(tidyr)

source("../peekbank-shiny/helpers/rt_helper.R")

theme_set(theme_classic() + 
            theme(text = element_text(size = 20))) 

```

## Developmental changes in gaze switching between target and distractor

See notes here:
<https://docs.google.com/document/d/1ySCNzheHPT261JNX44VyonfwkapzK212TEMS3DFS4xA/edit>

Next steps: [updated 08/28/2024]
- connect number of switches and fixation duration to overall accuracy and RT
- add image properties

```{r loading_data}
# Loading peekbank data into a variable called aoi_data_joined
load(here("cached_intermediates/1_aoi_data_joined.Rds"))
```
We select only datasets that were hand-coded (to keep the definition of switches a bit more consistent).
In the future we may want to find an equivalent for eye-tracking datasets
```{r selecting_datasets_for_analysis}

# select adams_marchman dataset (in the future this will be all hand-coded datasets)
hand_coded_df <- aoi_data_joined %>%
  filter(coding_method == "manual gaze coding")
```

```{r descriptives}

# Plot age distribution
hist(hand_coded_df$age)

```

Next, we tag all fixations with their start and end points, and
mark the 3 stages of the trial - before the critical event ("pre-critical"),
the first half of the trial after the critical event ("first half"),
and the second half of the trial after the critical event ("second half")
```{r tagging_switches}
# mark fixations to target, distractor, missing or other (coded in the aoi variable)
rle_data <- hand_coded_df %>%
  filter(any(t_norm == 0)) %>% # only use trials that have data at t==0
  group_by(dataset_name, dataset_id, subject_id,
           administration_id, trial_id) %>%
  reframe(min_tnorm=min(t_norm),# compute the first time point in the trial
          max_tnorm=max(t_norm), # compute the last time point in the trial
    lengths = rle(aoi)$lengths, # get the length of the fixation
            values = rle(aoi)$values) # get the object of the fixation

# add the onset and offset of fixations, as well as the stage
# of the trial that the fixation ended in (meaning, where the switch occurred)
frames_per_second = 25 # this is a fixed number for Peekbank
rle_data_with_times <-
  rle_data %>%
  group_by(dataset_name, dataset_id, subject_id,
           administration_id, trial_id) %>%
  mutate(
    total_trial_length=max_tnorm-min_tnorm, # get the duration of the trial
    cumulative_length_start = lag(cumsum(lengths),default=0), # onset of the fixation 
    cumulative_length_end = cumsum(lengths), # offset of the fixation
    t_start = min_tnorm+cumulative_length_start*frames_per_second, # convert to frames
    t_end = min_tnorm+(cumulative_length_end-1)*frames_per_second,# convert to frames
    # marking the stage of the trial as:
    # "pre-critical": from the onset of the trial, till the critical event
    # "first half": from the critical event, till the halfway point of the post-critical event trial
    # "second half": from the halfway point of the post-critical event trial till the end of the trial
    # these are marked based on the end time of the fixation, since that's when a switch occured
    trial_stage = ifelse(t_end < 0, "pre-critical",
                         ifelse(t_end < max_tnorm/2, "first half",
                                "second half")))

#join back in relevant data 
rle_data_with_times <-
  left_join(rle_data_with_times,
            hand_coded_df %>%
              select(subject_id, administration_id, trial_id,
                     age, dataset_name,
                     vanilla_trial,target_label, target_side,
                     stimulus_novelty, distractor_novelty,distractor_label) %>%
              distinct())
```

Next, we are computing the rate of switching, defined as the number of switches
divided by the duration of the interval of interest (in this case, the corresponding
stage of the trial)
```{r computing_switching_rate}

# compute the switching rate
rle_data_with_times <-
  rle_data_with_times %>%
  group_by(subject_id, trial_id, administration_id, trial_stage) %>%
  mutate(
    switch_num=length(values)-1, # number of trials per trial stage
    switch_num_per_trial_len = switch_num/total_trial_length,
    # get switching rate by dividing the number of switches by the duration of the trial stage
    length_ms=lengths*frames_per_second # ask Martin! 
  )

# create summary df for number of switches
rle_data_with_times_switchnum_summary =
  rle_data_with_times %>%
  # group data by subject, administration session, age, and trial stage
  group_by(administration_id, subject_id, age, trial_stage) %>%
  # get mean and SDs for number of switches, standardized by trial length
  summarise(mean = mean(switch_num_per_trial_len),
            sd = sd(switch_num_per_trial_len))

# create summary df for length of switches
rle_data_with_times_switchlength_summary =
  rle_data_with_times %>%
  # group data by subject, administration session, age, and trial stage
  group_by(administration_id, subject_id, age, trial_stage) %>%
  # get mean and SDs for LENGTH of switches
  summarise(mean = mean(length_ms),
            sd = sd(length_ms))


```
Next, we tag different types of switches. Specifically, we're tagging switches from 
target or distractor into missing data (or other), as well as switches between 
target and distractor
```{r computing_proportion_of_switches_by_type}

# getting the next fixation, so we can define each switch as a specific type 
rle_data_with_times <- rle_data_with_times %>%
          group_by(trial_id) %>%
          mutate(values_next = lead(values, order_by = trial_id)) %>%
          ungroup()

# Tag switch types:
# distractor -> missing/other
# target -> missing/other
# target -> distractor
# distractor -> target

switch_types <- rle_data_with_times %>%
  group_by(trial_id,trial_stage, subject_id,age, administration_id) %>%
  summarize(p_distractor2missing = mean(values == "distractor" & values_next %in% c("other","missing")),
            p_target2missing = mean(values == "target" & values_next %in% c("other","missing")),
            p_target2distractor = mean(values == "target" & values_next == "distractor"),
            p_distractor2target = mean(values == "distractor" & values_next == "target")) %>%
  ungroup()

# Reorganize to long format, and compute average proportions of switches in each
# category for each subject/administration by trial stage and switch type
switch_types_summary <- switch_types %>%
                      filter(trial_stage %in% c("first half", "second half")) %>%
                      pivot_longer(
                              cols = c(p_distractor2missing, p_target2missing,p_target2distractor,
                                       p_distractor2target),
                              names_to = "switch_type",
                              values_to = "proportion_of_switch_type")%>%
                      group_by(subject_id, administration_id, age,switch_type, trial_stage) %>%
                      summarize(mean_proportion = mean(proportion_of_switch_type, na.rm = TRUE)) %>%
                      ungroup()

# Plotting changes in the proportion for each switch type with age
ggplot(data = switch_types_summary %>% filter(switch_type %in% c('p_distractor2target','p_target2distractor')),
       aes(x = age,
           y = mean_proportion,
           fill = switch_type,
           color = switch_type)) +
  facet_grid(rows = vars(trial_stage))+
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  #scale_x_continuous(#limits = c(6, 84),
  #breaks = seq(6, 84, 6)) +
  #scale_y_continuous(limits = c(0, .007),
  #breaks = seq(0, .007, .001)) +
  labs(y = "Mean # of Switches (Std. by Trial Length)",
     x = "Age in months",
     fill = "Trial Stage",
     color = "Trial Stage") +
  theme_minimal()

# regression looking at switches between target and distractor
summary(lm(mean_proportion ~ switch_type*age,switch_types_summary%>% filter(switch_type %in% c('p_distractor2target','p_target2distractor'))))



# plot looking at switches to missing/other
ggplot(data = switch_types_summary %>% filter(switch_type %in% c('p_distractor2missing','p_target2missing')),
       aes(x = age,
           y = mean_proportion,
           fill = switch_type,
           color = switch_type)) +
  facet_grid(rows = vars(trial_stage))+
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  #scale_x_continuous(#limits = c(6, 84),
  #breaks = seq(6, 84, 6)) +
  #scale_y_continuous(limits = c(0, .007),
  #breaks = seq(0, .007, .001)) +
  labs(y = "Mean # of Switches (Std. by Trial Length)",
     x = "Age in months",
     fill = "Trial Stage",
     color = "Trial Stage") +
  theme_minimal()
# regression looking at switches to missing/other
summary(lm(mean_proportion ~ switch_type*age,switch_types_summary%>% filter(switch_type %in% c('p_distractor2missing','p_target2missing'))))
```

Plotting number of switches by age



```{r n_switches_by_age}

# number of switches by age and trial_stage
ggplot(data = rle_data_with_times_switchnum_summary,
       aes(x = age,
           y = mean,
           fill = trial_stage,
           color = trial_stage)) +
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  #scale_x_continuous(#limits = c(6, 84),
  #breaks = seq(6, 84, 6)) +
  #scale_y_continuous(limits = c(0, .007),
  #breaks = seq(0, .007, .001)) +
  labs(y = "Mean # of Switches (Std. by Trial Length)",
     x = "Age in months",
     fill = "Trial Stage",
     color = "Trial Stage") +
  theme_minimal()

# regression looking at number of switches by age for each trial stage
summary(lm(mean ~ age,rle_data_with_times_switchnum_summary%>% filter(trial_stage %in% c('first half'))))
summary(lm(mean ~ age,rle_data_with_times_switchnum_summary%>% filter(trial_stage %in% c('second half'))))
summary(lm(mean ~ age,rle_data_with_times_switchnum_summary%>% filter(trial_stage %in% c('pre-critical'))))

```
Plotting durations of fixations by age
```{r fixation_durations_by_age}

# length of switches by age and trial_stage
ggplot(data = rle_data_with_times_switchlength_summary,
       aes(x = age,
           y = mean,
           fill = trial_stage,
           color = trial_stage)) +
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  #scale_x_continuous(limits = c(6, 84),
  #                   breaks = seq(6, 84, 6)) +
  #scale_y_continuous(limits = c(0, 11000),
  #                   breaks = seq(0, 11000, 2750)) +
  labs(y = "Mean Length of Fixations",
       x = "Age in months",
       fill = "Trial Stage",
       color = "Trial Stage") +
  theme_minimal()


# regression looking at duration of fixations by age for each trial stage
summary(lm(mean ~ age,rle_data_with_times_switchlength_summary%>% filter(trial_stage %in% c('first half'))))
summary(lm(mean ~ age,rle_data_with_times_switchlength_summary%>% filter(trial_stage %in% c('second half'))))
summary(lm(mean ~ age,rle_data_with_times_switchlength_summary%>% filter(trial_stage %in% c('pre-critical'))))

```

------------------------------------------------------------------------------------------

MARTINS CODE

Average switches per administration

```{r}
# admin_rle_data_with_times <- rle_data_with_times %>%
#   group_by(dataset_name, dataset_id,subject_id, administration_id,age,
#            vanilla_trial,target_label, target_side,
#            stimulus_novelty, distractor_novelty,distractor_label) %>%
#   summarize(
#     avg_switch_num = mean(switch_num,na.rm=T),
#     avg_switch_norm = mean(switch_num_per_trial_len),
#     avg_length=mean(length_ms),
#     sd_length=sd(length_ms)
#   )

# ggplot(admin_rle_data_with_times,aes(age,avg_switch_num)) +
#   geom_point(alpha=0.1)+
#   geom_smooth(method="lm")#+
#   #facet_wrap(~dataset_name)
# 
# ggplot(admin_rle_data_with_times,aes(age,avg_length)) +
#   geom_point(alpha=0.1)+
#   geom_smooth(method="loess",se=F)#+
#   #facet_wrap(~dataset_name)
```

