---
title: "longitudinal_switching_analysis"
output: html_document
date: '2024-08-27'
---

```{r setup, include=FALSE}

# load relevant libraries
knitr::opts_chunk$set(echo = TRUE)
library(here)         # for wd management
library(data.table)   # for data display
library(ggplot2)      # for plotting
library(dplyr)        # for everything else
library(tidyr)

# load helper file that computes RTs
source("../peekbank-shiny/helpers/rt_helper.R")

# set standard plotting theme
theme_set(theme_classic() + 
            theme(text = element_text(size = 20))) 

```

## Developmental changes in gaze switching between target and distractor

See notes here:
<https://docs.google.com/document/d/1ySCNzheHPT261JNX44VyonfwkapzK212TEMS3DFS4xA/edit>

Next steps: [updated 08/28/2024] - connect number of switches and fixation duration to
overall accuracy and RT - add image properties

### Setup

```{r loading_data}

# Loading peekbank data into a variable called aoi_data_joined
load(here("cached_intermediates/1_aoi_data_joined.Rds"))
load(here("../peekbank-method/cached_intermediates/1_d_trial.Rds"))

```

### Data Import and Cleaning

We select only datasets that were hand-coded (to keep the definition of switches a bit
more consistent). In the future we may want to find an equivalent for eye-tracking
datasets

```{r selecting_datasets_for_analysis}

# select all hand-coded data sets (tagged as "manual gaze coding")
hand_coded_df <- aoi_data_joined %>%
  filter(coding_method == "manual gaze coding")

Hmisc::describe(hand_coded_df$dataset_name) # k  = 17 data sets
Hmisc::describe(hand_coded_df$subject_id) # N = 906 subjects

```

### Descriptives

```{r descriptives}

# plot age distribution
hand_coded_df_age =
  hand_coded_df %>%
  # collapse so that there is only one row per subject
       group_by(subject_id) %>%
  # grab age per subject
       summarise(age = mean(age))

hist(hand_coded_df_age$age)

```

### Data Tagging: Trial Stages

Next, we tag all fixations with their start and end points, and mark the 3 stages of the
trial - before the critical event ("pre-critical"), the first half of the trial after the
critical event ("first half"), and the second half of the trial after the critical event
("second half")

```{r tagging_switches}
# mark fixations to target, distractor, missing or other (coded in the aoi variable)
rle_data <- hand_coded_df %>%
  filter(any(t_norm == 0)) %>% # only use trials that have data at t==0
  group_by(dataset_name, dataset_id, subject_id,
           administration_id, trial_id) %>%
  reframe(min_tnorm=min(t_norm),# compute the first time point in the trial
          max_tnorm=max(t_norm), # compute the last time point in the trial
    lengths = rle(aoi)$lengths, # get the length of the fixation
            values = rle(aoi)$values) # get the object of the fixation

# add the onset and offset of fixations, as well as the stage
# of the trial that the fixation ended in (meaning, where the switch occurred)
frames_per_second = 25 # this is a fixed number for Peekbank
rle_data_with_times <-
  rle_data %>%
  group_by(dataset_name, dataset_id, subject_id,
           administration_id, trial_id) %>%
  mutate(
    total_trial_length=max_tnorm-min_tnorm, # get the duration of the trial
    cumulative_length_start = lag(cumsum(lengths),default=0), # onset of the fixation 
    cumulative_length_end = cumsum(lengths), # offset of the fixation
    t_start = min_tnorm+cumulative_length_start*frames_per_second, # convert to frames
    t_end = min_tnorm+(cumulative_length_end-1)*frames_per_second,# convert to frames
    # marking the stage of the trial as:
    # "pre-critical": from the onset of the trial, till the critical event
    # "first half": from the critical event, till the halfway point of the post-critical event trial
    # "second half": from the halfway point of the post-critical event trial till the end of the trial
    # these are marked based on the end time of the fixation, since that's when a switch occured
    trial_stage = ifelse(t_end < 0, "pre-critical",
                         ifelse(t_end < max_tnorm/2, "first half",
                                "second half")))

#join back in relevant data 
rle_data_with_times <-
  left_join(rle_data_with_times,
            hand_coded_df %>%
              select(subject_id, administration_id, trial_id,
                     age, dataset_name,
                     vanilla_trial,target_label, target_side,
                     stimulus_novelty, distractor_novelty,distractor_label) %>%
              distinct())
```

### Compute Rate of Switching

Next, we are computing the rate of switching, defined as the number of switches divided by
the duration of the interval of interest (in this case, the corresponding stage of the
trial)

```{r computing_switching_rate}

# compute the switching rate
rle_data_with_times <-
  rle_data_with_times %>%
  group_by(subject_id, trial_id, administration_id, trial_stage) %>%
  mutate(
    switch_num=length(values)-1, # number of trials per trial stage
    switch_num_per_trial_len = switch_num/total_trial_length,
    # get switching rate by dividing the number of switches by the duration of the trial stage
    length_ms=lengths*frames_per_second # ask Martin! 
  )

# create summary df for number of switches
rle_data_with_times_switchnum_summary =
  rle_data_with_times %>%
  # group data by subject, administration session, age, and trial stage
  group_by(administration_id, subject_id, age, trial_stage) %>%
  # get mean and SDs for number of switches, standardized by trial length
  summarise(mean = mean(switch_num_per_trial_len),
            sd = sd(switch_num_per_trial_len))


# create summary df for number of switches by trial type
rle_data_with_times_switchnum_by_trial_type_summary =
  rle_data_with_times %>%
  # group data by subject, administration session, age, and trial stage
  group_by(administration_id, subject_id, trial_stage,stimulus_novelty) %>%
  # get mean and SDs for number of switches, standardized by trial length
  summarise(mean = mean(switch_num_per_trial_len),
            sd = sd(switch_num_per_trial_len))

# create summary df for length of switches
rle_data_with_times_switchlength_summary =
  rle_data_with_times %>%
  # group data by subject, administration session, age, and trial stage
  group_by(administration_id, subject_id, age, trial_stage) %>%
  # get mean and SDs for LENGTH of switches
  summarise(mean = mean(length_ms),
            sd = sd(length_ms))

```

### Compute Accuracy and RT

```{r computing_ccuracy_and_RT}
  rts_by_trial <- rle_data %>%
                      group_by(administration_id, trial_id, subject_id) %>%
                      nest() %>%
                      mutate(data = lapply(data, get_rt)) %>%
                      unnest(cols = c(data)) %>%
                      left_join(d_trial %>%
                                  select(-t_norm, -correct, -aoi) %>%
                                  distinct())

    rt_stats_by_participant <- rts_by_trial %>% 
            filter(shift_start_rt > 300)%>%
              group_by(administration_id, subject_id) %>%
              summarise(nas = mean(is.na(shift_start_rt)), 
                        too_fast = mean(shift_start_rt < 300, na.rm=TRUE), 
                        d_t = mean(shift_type == "D-T", na.rm=TRUE), 
                        t_d = mean(shift_type == "T-D", na.rm=TRUE),
                        other = mean(shift_type == "other", na.rm=TRUE),
                        no_shift = mean(shift_type == "no shift", na.rm=TRUE))


  
    accuracy_by_trial <- d_trial |>
                  filter(t_norm > 200, t_norm < 2000) |>
                  group_by(administration_id, trial_id) |>
                  summarise(accuracy = mean(correct, na.rm=TRUE),
                            prop_data = mean(!is.na(correct)))
    
    
    accuracy_stats_by_participant <- accuracy_by_trial %>% 
              group_by(administration_id) %>%
              summarise(avg_accuracy = mean(accuracy, na.rm=TRUE))
    
    
 
  
```

### Analyses

#### Number of Switches \~ Age (By Trial Stage)

Plotting number of switches by age Summary: No developmental change in pre-critical event
switches, but a decrease in switches during the post-critical event parts of the trial
(both first and second half)

```{r n_switches_by_age}

# number of switches by age and trial_stage
ggplot(data = rle_data_with_times_switchnum_summary,
       aes(x = age,
           y = mean,
           fill = trial_stage,
           color = trial_stage)) +
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  labs(y = "Mean # of Switches (Std. by Trial Length)",
     x = "Age in months",
     fill = "Trial Stage",
     color = "Trial Stage") +
  theme_minimal()

# regression looking at number of switches by age for each trial stage
summary(lm(mean ~ age,rle_data_with_times_switchnum_summary%>% filter(trial_stage %in% c('first half'))))
summary(lm(mean ~ age,rle_data_with_times_switchnum_summary%>% filter(trial_stage %in% c('second half'))))
summary(lm(mean ~ age,rle_data_with_times_switchnum_summary%>% filter(trial_stage %in% c('pre-critical'))))



```

#### Number of Switches \~ Stimulus Familiarity (By Trial Stage)

Number of switches by stimulus familiarity Summary: There are fewer switches for familiar
(compared to novel) stimuli:

```{r switch_number_by_trial_type}
# number of switches by age and trial_stage
rle_data_with_times_switchnum_by_trial_type_summary$stimulus_novelty = as.factor(rle_data_with_times_switchnum_by_trial_type_summary$stimulus_novelty)

rle_data_with_times_switchnum_by_trial_type_summary$trial_stage = factor(rle_data_with_times_switchnum_by_trial_type_summary$trial_stage, levels = c('pre-critical','first half','second half'))

ggplot(data = rle_data_with_times_switchnum_by_trial_type_summary,
       aes(x = trial_stage,
           y = mean,
           fill = stimulus_novelty,
           color = stimulus_novelty)) +
  geom_boxplot() +
  labs(y = "Mean # of Switches (Std. by Trial Length)",
     x = "Trial Stage",
     fill = "Vanilla Trial",
     color = "Vanilla Trial") +
  theme_minimal()


summary(lm(mean ~ stimulus_novelty, data = rle_data_with_times_switchnum_by_trial_type_summary))

```

#### Duration of Fixation \~ Age (By Trial Stage)

Duration of fixations by age Summary: The duration of fixations increases with age for all
stages of the trial, but especially the last half of the post-critical event

```{r fixation_durations_by_age}

# length of switches by age and trial_stage
ggplot(data = rle_data_with_times_switchlength_summary,
       aes(x = age,
           y = mean,
           fill = trial_stage,
           color = trial_stage)) +
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  labs(y = "Mean Length of Fixations",
       x = "Age in months",
       fill = "Trial Stage",
       color = "Trial Stage") +
  theme_minimal()



# regression looking at duration of fixations by age for each trial stage
summary(lm(mean ~ age,rle_data_with_times_switchlength_summary%>% filter(trial_stage %in% c('first half'))))
summary(lm(mean ~ age,rle_data_with_times_switchlength_summary%>% filter(trial_stage %in% c('second half'))))
summary(lm(mean ~ age,rle_data_with_times_switchlength_summary%>% filter(trial_stage %in% c('pre-critical'))))

```

#### Switches Between Target & Distractor \~ Accuracy (Only For Pre-Critical)

Looking at how switching between target and distractor relates to accuracy in the trial
Kind of confusing skewed data, so a bit tricky to interpret?

```{r accuracy_and_RT_by_number_of_pretrial_switches}
  RT_and_switching_by_participant <-
                  left_join(rle_data_with_times_switchnum_summary,
                            rt_stats_by_participant %>%
                              select(administration_id, subject_id,
                                     avg_RT = d_t) %>%
                              distinct())




# length of switches by age and trial_stage
ggplot(data = RT_and_switching_by_participant,
       aes(x = mean,
           y = avg_RT)) +
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  labs(y = "Avegare RT (participant)",
       x = "Switch rate for pre-critical event part of the trial") +
  theme_minimal()
summary(lm(avg_RT~mean+age,RT_and_switching_by_participant%>%filter(trial_stage == 'pre-critical')))

  accuracy_and_switching_by_participant <-
                  left_join(rle_data_with_times_switchnum_summary,
                            accuracy_stats_by_participant %>%
                              select(administration_id,avg_accuracy) %>%
                              distinct())
  
  # length of switches by age and trial_stage
ggplot(data = accuracy_and_switching_by_participant,
       aes(x = mean,
           y = avg_accuracy)) +
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  labs(y = "Avegare accuracy (participant)",
       x = "Switch rate for pre-critical event part of the trial") +
  theme_minimal()

summary(lm(avg_accuracy~mean+age,accuracy_and_switching_by_participant%>%filter(trial_stage == 'pre-critical')))

```

### Data Tagging: Different Types of Switches

Next, we tag different types of switches. Specifically, we're tagging switches from target
or distractor into missing data (or other), as well as switches between target and
distractor

```{r computing_proportion_of_switches_by_type}

# getting the next fixation, so we can define each switch as a specific type 
rle_data_with_times <- rle_data_with_times %>%
          group_by(trial_id) %>%
          mutate(values_next = lead(values, order_by = trial_id)) %>%
          ungroup()

# Tag switch types:
# distractor -> missing/other
# target -> missing/other
# target -> distractor
# distractor -> target

switch_types <- rle_data_with_times %>%
  group_by(trial_id,trial_stage, subject_id,age, administration_id) %>%
  summarize(p_distractor2missing = mean(values == "distractor" & values_next %in% c("other","missing")),
            p_target2missing = mean(values == "target" & values_next %in% c("other","missing")),
            p_target2distractor = mean(values == "target" & values_next == "distractor"),
            p_distractor2target = mean(values == "distractor" & values_next == "target")) %>%
  ungroup()

# Reorganize to long format, and compute average proportions of switches in each
# category for each subject/administration by trial stage and switch type
switch_types_summary <- switch_types %>%
                      filter(trial_stage %in% c("first half", "second half")) %>%
                      pivot_longer(
                              cols = c(p_distractor2missing, p_target2missing,p_target2distractor,
                                       p_distractor2target),
                              names_to = "switch_type",
                              values_to = "proportion_of_switch_type")%>%
                      group_by(subject_id, administration_id, age,switch_type, trial_stage) %>%
                      summarize(mean_proportion = mean(proportion_of_switch_type, na.rm = TRUE)) %>%
                      ungroup()

```

### Analyses

#### Switches to Missing/Other \~ Age

```{r}

# plot looking at switches to missing/other
ggplot(data = switch_types_summary %>% filter(switch_type %in% c('p_distractor2missing','p_target2missing')),
       aes(x = age,
           y = mean_proportion,
           fill = switch_type,
           color = switch_type)) +
  facet_grid(rows = vars(trial_stage))+
  geom_point(alpha = .05) +
  geom_smooth(method = 'lm') +
  labs(y = "Mean # of Switches (Std. by Trial Length)",
     x = "Age in months",
     fill = "Trial Stage",
     color = "Trial Stage") +
  theme_minimal()
# regression looking at switches to missing/other
summary(lm(mean_proportion ~ switch_type*age,switch_types_summary%>% filter(switch_type %in% c('p_distractor2missing','p_target2missing'))))

```

------------------------------------------------------------------------------------------

MARTINS CODE

Average switches per administration

```{r}
# admin_rle_data_with_times <- rle_data_with_times %>%
#   group_by(dataset_name, dataset_id,subject_id, administration_id,age,
#            vanilla_trial,target_label, target_side,
#            stimulus_novelty, distractor_novelty,distractor_label) %>%
#   summarize(
#     avg_switch_num = mean(switch_num,na.rm=T),
#     avg_switch_norm = mean(switch_num_per_trial_len),
#     avg_length=mean(length_ms),
#     sd_length=sd(length_ms)
#   )

# ggplot(admin_rle_data_with_times,aes(age,avg_switch_num)) +
#   geom_point(alpha=0.1)+
#   geom_smooth(method="lm")#+
#   #facet_wrap(~dataset_name)
# 
# ggplot(admin_rle_data_with_times,aes(age,avg_length)) +
#   geom_point(alpha=0.1)+
#   geom_smooth(method="loess",se=F)#+
#   #facet_wrap(~dataset_name)
```
